Making Sequential Programs Deterministic

### 2019-04-01 design.tex
Edits in sections 2-2.3

\subsection{Preliminaries}

@@

In a type system,
a \textit{type} represents a set of values.  It abstracts or restricts the
set of possible run-time values that an expression may evaluate to.
A programming language provides \emph{basetypes}, such as \<int>.
A \textit{type qualifier} on a basetype adds additional constraints;
that is, it reduces the size of the set of values.
An example type qualifier is \<Positive>,\todo{should all type qualifiers
  be lowercase in this section??} and a type (which combines a qualifier
and a basetype) is \<Positive int>.
A polymorphic type such as \<List> can be instantiated by a type argument,
as in \ptype{List}{Positive int}.

A type checker verifies the types written in a program.

@@

    \caption{A sampling of standard typing rules.  We omit other standard rules,
    for brevity.  $\sqsubseteq$ is overloaded for classes and types.}
    \label{typecheck-rules-standard}
\end{figure}

    \caption{Subtyping rules for \ourTypeSystem.  $\sqsubseteq$ is overloaded for classes,
    basetypes, type qualifiers, and types.}
    \label{fig:typecheck-rules}
\end{figure}

@@

Figure~\cref{typecheck-rules-standard} shows a sample of the standard typing rules applied to 
any typed language. In the generic case, the type $\tau$ represents a valid basetype in a programming language.
\ourTypeSystem extends a basetype with a type qualifier. Throughout this paper, $\tau$ represents a fully qualified type.
In other words, type $\tau = \kappa \ \beta$ consists of both a qualifier $\kappa$ and a basetype $\beta$, neither of which may be omitted.
Further, $\beta$ is a type without a type qualifier, but which has type qualifiers on all its type parameters.
Formally, $\beta ::= C|C \angles{\tau}$ where $C$ represents a valid Java basetype.
\ourTypeSystem checks all the standard typing rules including the ones shown in Figure~\cref{typecheck-rules-standard} with
the only difference of having an additional type qualifier on each of the basetypes. The type qualifiers can only 
constrain the set of valid types. As a result, \ourTypeSystem doesn't allow any behavior that the original type system doesn't.

@@

The core of the determinism type system is the following type qualifiers:
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map that contains the same elements in every execution, but possibly
      in a different order.
    \item \<Det> indicates that
    the expression evaluates to the same value (with respect to \<.equals()>) in all
    executions; for a collection, iteration also yields the values in the same
    order.
\end{itemize}
\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show their subtyping relationship.

@@

@@

The element type of a \<NonDet> collection must be \<NonDet>.
Otherwise, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection}\ rules of
\cref{type-validity}). Note that the \rulename{ordernondet} versions also handle \<Det>,
because $\<Det> \sqsubseteq \<OrderNonDet>$.
\Cref{fig:determinism-collections} restates its restrictions, and here are examples.

\smallskip
\noindent
\begin{minipage}{.48\textwidth}
Some examples of valid types are:
% Using "~" instead of "\ " does not work within \codeid :-(
\begin{smaller}
\begin{Verbatim}[commandchars=\\\{\}]
NonDet      \lptype{List}{NonDet      Integer}
OrderNonDet \lptype{List}{OrderNonDet \lptype{Set}{...}}
OrderNonDet \lptype{List}{Det         Integer}
Det         \lptype{List}{Det         Integer}

\end{Verbatim}
\end{smaller}
\end{minipage}
\hfill
\begin{minipage}{.46\textwidth}
These types are invalid:
\begin{smaller}
\begin{Verbatim}[commandchars=\\\{\}]
NonDet      \lptype{List}{OrderNonDet \lptype{Set}{...}}
NonDet      \lptype{List}{Det         Integer}
OrderNonDet \lptype{List}{NonDet      Integer}
Det         \lptype{List}{NonDet      Integer}
Det         \lptype{List}{OrderNonDet \lptype{Set}{...}}
\end{Verbatim}
\end{smaller}
\end{minipage}
\smallskip

 Although \<NonDet> collections containing \<Det> or \<OrderNonDet> elements could exist, they are
disallowed to prevent the following bug:

\begin{verbatim}
public static void add(@NonDet List<@Det String> list, @NonDet int index, @Det String s) {
    list.add(index, s);
}

public static void f(@Det List<@Det String> list, @NonDet int index, @Det String s) {
    add(list, index, s);
}
\end{verbatim}

The possibility of mutation allows us to add to the \<Det List> at a
\<NonDet> index, which is unsound.

\todo{Some notes:}
Typical type for \<get> is \\
$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$ \\
but we need to change that.  If either argument is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.
If both arguments are \<Det>, the result is $\tau$ as usual.
\todo{In that case, we know that $\tau = \<Det> \beta$.}

A collection or map annotated as \<OrderNonDet> has the following properties:
%\todo{Write these as type rules.}
\begin{enumerate}
    \item The individual elements extracted by iterating over the collection get the type \<NonDet> (Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules}). 
    Rule \rulename{hasnext element} states that querying if the iterator of an \<OrderNonDet Collection> \<hasNext()> will result in a \<Det boolean>.
    \item Accessing properties such as \<size()> or querying if the collection \<isEmpty()> will return types
    that are annotated as \<Det> as they do not depend on the iteration order. 
\end{enumerate}

>>>>

\subsection{Preliminaries and notation}

@@

\begin{figure}
  \newcommand{\bnfalt}{\ \ |\ \ }
  \begin{tabular}{lcll}
    $C$ & ::= & \<Int>\bnfalt \<List>\bnfalt \ldots & class name \\
    $\kappa$ & ::= & \<NonDet>\bnfalt \<OrderNonDet>\bnfalt\<Det> & type qualifier \\
    $\tau$ & ::= & $\kappa\ \beta$ & \mbox{type} \\
    $\beta$ & ::= & $C$\bnfalt $C\angles{\overline{\tau}}$ & \mbox{unqualified type}
  \end{tabular}
  \caption{Grammar of types}
  \label{fig:grammar}
\end{figure}

In a type system,
a \textit{type} represents a set of values.  It abstracts or restricts the
set of possible run-time values that an expression may evaluate to.
A programming language provides \emph{basetypes}, such as \<Int>.
A \textit{type qualifier} on a basetype adds additional constraints;
that is, it reduces the size of the set of values.
An example type qualifier is \<Positive>, and a type (which combines a qualifier
and a basetype) is \<Positive Int>.
A polymorphic type abstraction such as \<List> can be instantiated by a type argument,
as in \ptype{List}{Positive Int}.

\Cref{fig:grammar} formalizes these notions.
\Cref{fig:grammar}'s $\beta$ is not a basetype as described above:  $\beta$
lacks a top-level type, but any type arguments are full types
$\tau$.  This paper does not formally define basetype, and uses $\beta$,
for simplicity of exposition.
In our core calculus, a type $\tau$ always has a qualifier.
Our implementation, \theDeterminismChecker, uses inference and defaults to
permit users to omit type qualifiers\todo{Forward reference}.

A type checker verifies the types written in a program.
A type qualifier constrain the set of possible run-time values, that is,
$\kappa \ /beta \sqsubseteq \beta$.
As a result, a qualifier type system does not allow any behavior that the
original type system does not.
However, the qualifier type system may reject more programs, and thus
affords stronger guarantees.
(Our implementation \theDeterminismChecker runs the underlying base type
system, then issues additional errors and warnings.)
\todo{Some qualifier type systems can be defined independently of the underlying
type system.  We don't do that because there are interactions between the
basetypes and the type qualifiers.  This allows greater precision, which
is important in practice.}

    \caption{Subtyping rules for \ourTypeSystem.  $\sqsubseteq$ is overloaded for classes,
    type qualifiers, types, and unqualified types $\beta$.}
    \label{fig:typecheck-rules}
\end{figure}

@@

    \caption{A sampling of standard typing rules.  We omit other standard rules,
    for brevity.}
    \label{typecheck-rules-standard}
\end{figure}

@@

\Cref{typecheck-rules-standard} shows a sample of standard typing rules for
an object-oriented programming language.

\OurTypeSystem checks all the standard typing rules including the ones shown in \cref{typecheck-rules-standard} with
the only difference of having an additional type qualifier on each of the basetypes.

@@

The core of the determinism type system is the following type qualifiers:
%\todo{introduce just @Det and @NonDet first, and then @OrderNonDet after polymorphism.}
\begin{itemize}
    \item \<NonDet> indicates
    that the expression might have different values in two different executions.
    \item \<OrderNonDet> indicates that the expression is a collection or
      map that contains the same elements in every execution, but possibly
      in a different order.
    \item \<Det> indicates that the expression evaluates to equal values in
      all executions; for an unordered collection such as a set, iteration
      also yields the values in the same order.
\end{itemize}
\Cref{fig:typecheck-rules,fig:determinism-hierarchy} show their subtyping relationship.

@@

\subsubsection{Type well-formedness}

@@

The element type of a \<NonDet> collection must be \<NonDet>.
Otherwise, the type qualifier on a collection's element must be a subtype or equal to
the collection's type qualifier (\rulename{collection} rules of
\cref{type-validity}). Note that the \rulename{ordernondet} versions also handle \<Det>,
because $\<Det> \sqsubseteq \<OrderNonDet>$.
\Cref{fig:determinism-collections} restates the \rulename{collection} rules
of
\cref{type-validity}.

\smallskip
\noindent
\begin{minipage}{.48\textwidth}
Some examples of valid types are:
% Using "~" instead of "\ " does not work within \codeid :-(
\begin{Verbatim}[commandchars=\\\{\}]
NonDet      \lptype{List}{NonDet      Int}
OrderNonDet \lptype{List}{OrderNonDet \lptype{Set}{...}}
OrderNonDet \lptype{List}{Det         Int}
Det         \lptype{List}{Det         Int}

\end{Verbatim}
\end{minipage}
\hfill
\begin{minipage}{.46\textwidth}
These types are invalid:
\begin{smaller}
\begin{Verbatim}[commandchars=\\\{\}]
NonDet      \lptype{List}{OrderNonDet \lptype{Set}{...}}
NonDet      \lptype{List}{Det         Int}
OrderNonDet \lptype{List}{NonDet      Int}
Det         \lptype{List}{NonDet      Int}
Det         \lptype{List}{OrderNonDet \lptype{Set}{...}}
\end{Verbatim}
\end{smaller}
\end{minipage}
\smallskip

It is possible to store an expression of type \<Det> in a \<NonDet>
collection, but all elements retrieved from a \<NonDet> collection have
type \<NonDet>.  If the type \ptype{NonDet List}{Det String} were
permitted, the following type hole would exist:

\begin{Verbatim}[commandchars=\\\{\}]
\lptype{Det    List}{Det String} ddlist
\lptype{NonDet List}{Det String} ndlist = ddlist   // permitted by subtyping rules
NonDet Int ni = ...
ndlist[ni] = "anywhere"
\end{Verbatim}

\noindent
The deterministic string \<"anywhere"> is placed in the list at an
arbitrary (that is, nondeterministic) index.  This must be prohibited,
because it unsoundly permits the deterministic list \<ddlist> to differ
from execution to execution.

\subsubsection{Behavior of order-nondeterministic collections}

A collection of type \<OrderNonDet> has special properties, including the following

\begin{enumerate}
\item
The individual elements retrieved from it have type \<NonDet>.  This
affects access, iteration, searching, and more.
(Rules \rulename{iteration} and \rulename{next element} in Figure~\ref{fig-ordernondet-rules})
\item
Size-related operations return a deterministic result.  This also affects
queries of whether an iterator has more elements.
\item
If the collection is sorted, or its elements are placed in a collection
that does sorting, the result is deterministic.
\end{enumerate}

To state the first point more formally, the 
typical type for list access, such as \<get>, is \\
$$\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau$$
but this type is incorrect for \ourTypeSystem.  Each of the following is a 
$$
\forall \kappa\ \beta . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau \\
\forall \tau . \<List>\angles{\tau} \times \<int> \rightarrow \tau
$$

but we need to change that.  If either argument is \<OrderNonDet> or \<NonDet>, then the
result is \<NonDet>.
If both arguments are \<Det>, the result is $\tau$ as usual.
\todo{In that case, we know that $\tau = \<Det> \beta$.}
